Seagrass mapping and monitoring \\ along the coasts of Crete, Greece

### 2011-02-12 MScDiplom_Lemenkova.tex
Valentijn's comments: 1, 2, 3, 4, 5, 7: Updated Hypotheses, Assumtions, Research Questions. Added long names of Chapters to ToC.

\documentclass[11pt]{article}
\special{papersize=140mm,210mm}
\usepackage{authblk}
\usepackage[english]{babel}
%\usepackage{authorindex}
\usepackage{color}
\usepackage[stable]{footmisc }
\usepackage{setspace}
\usepackage[printonlyused,withpage]{acronym}
\usepackage[left=30mm,right=20mm,top=30mm,bottom=35mm]{geometry}
\usepackage{pslatex}
%\usepackage{times}
%\usepackage{psnfss}
%\usepackage{glossary}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{calc}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage[svgnames,table]{xcolor}
\usepackage[subfigure]{tocloft}
%\usepackage[table]{xcolor}
\usepackage{paralist}
%\usepackage{makeidx}\makeindex
%\usepackage[makeindex,protected]{splitidx}
%\usepackage{index}
\usepackage{imakeidx}
\makeindex[name=concepts,title=Index of Concepts,intoc]
\makeindex[name=people,title=Index of People,intoc,columns=3,columnseprule]
\usepackage{fancyhdr}
\pagestyle{fancy}
\usepackage{fancyheadings}
\topmargin = 1pt
\headheight=10mm
\headsep=7pt
%\ensuremath
\usepackage{lastpage}
\usepackage{float}
\usepackage{subfig}
\usepackage{caption}
\usepackage{chngcntr}
\counterwithin{figure}{section}
\counterwithin{table}{section}
\usepackage{appendix}
\captionsetup{margin=10pt,font=small,labelfont=bf}
\usepackage{wrapfig}
\definecolor{darkblue}{rgb}{0.0,0.0,0.3}
\setcounter{tocdepth}{4}
\parindent 0pt
%\floatstyle{boxed} 
\restylefloat{figure}
\usepackage{hyperref}
\hypersetup{pdftitle={Seagrass mapping and monitoring along the coasts of Crete, Greece}, pdfauthor={Lemenkova Polina}, pdfsubject={MScThesis}, pdfcreator={Lemenkova Polina}, pdfproducer={Lemenkova Polina}, colorlinks=true,linkcolor=Navy, citecolor=Crimson, filecolor=magenta, urlcolor=blue,pdfkeywords={P.oceanica, seagrass, spectral reflectance, remote sensing, WASI, RTM, Google Earth, mapping, Crete}}

@@

\subsection{Research questions}\index[concepts]{Research!questions}
	\begin{itemize}	
		\item Is \textit{P.oceanica} spectrally distinct from other seafloor cover types with varying \textit{in-situ} conditions ?
		\item Can broadband and hyperspectral remote sensing data be used for the mapping of \textit{P.oceanica}: what sensor yields enough radiometric details to discriminate seagrass from other seafloor cover types ?
	\end{itemize}

\subsection{Hypotheses testing}\index[concepts]{Research!questions!hypothesis}
A statistical testing will be used to compare between the spectral responses of the different seafloor cover types (i.e. sand and \textit{P. oceanica}),
 whether it is spectrally distinct and at least one pair is statistically different at every spectral
band.
Hypothesis $H_o$: seagrass types are not spectrally distinct from other seafloor types with
varying \textit{in-situ} conditions, which means
$H_o$: μ1 =μ2 =μ3=...= μn.
The alternative Hypothesis $H_a$ claims the opposite statement: seagrass is spectrally
distinct with varying \textit{in-situ} conditions, $H_o$: μ$\neq$μ2$\neq$μ3$\neq$ ... $\neq$μn.
The distribution of the spectral responses at every spectral band is assumed to be normal as well as
the equality of the statistical variances.
The hypothesis testing is suggested to be carried out using the \ac{ANOVA} statistical test. The purpose of
\ac{ANOVA} test is to visualize in an effective and quick way the spectral differences between seagrass
species and their spatial distribution. Thus, the key hypotheses of the research will be tested to prove
whether the results of the research are accurate, reasonable and correct.

@@

The spectral reflectance, distinct for each seafloor cover type, enables their discrimination on the image. The pre-processing
of the images includes imagery corrections for atmospheric noises and effects of the water
column. Reflectance spectra of the seagrass canopy at different depths of the water-column are
analysed for the discrimination of their spectral signatures, enabling to separate various seafloor types
during classification. The results of the of imagery classification are analysed for the detection of the
dynamics in \textit{P.oceanica} seagrass distribution along the northern coasts of Crete. Aerial imagery from
Google Earth with high spatial resolution, suitable for the large-scale detailed mapping of seagrass
mattes, is used for the improvement of the accuracy of large seagrass meadows and separate mattes
within the meadows. The \textit{in situ} underwater videometric measurements of the seafloor are collected
during the fieldwork in Crete, for the validation of the classification results and to determine the exact
current distribution of the \textit{P.oceanica} meadows. The image processing includes steps of the remote
sensing techniques, i.e. calibration, masking from land and cloud, atmospheric correction, sea surface
glint and depth effects correction as recommended \cite{Matarrese08}\label{Matarrese08}. During the image classification
working step the training sites for the supervised classification methods are designed, as well as its
control and trials of different classification approaches (Unsupervised, K-means or Isodata;
Supervised, Maximum Likelihood).
\pagebreak

\subsection{Research assumptions}\index[concepts]{Research!assumptions}
The general research assumption that is used for further applications of the remote sensing methods of this work is that optical properties of the spectral reflectance of the seawater with different bottom substrata enable to discriminate seafloor cover types using remote sensing products from broadband sensors.\vspace{1ex}\\
The accepted model-specific parameters of the \ac{WASI} simulator are described in Table 4.2. The weather conditions for the measurements are assumed to be perfect: clear, sunny, windless days, because wind roughen water surface causing sun glitter, therefore, values of spectral reflectance may contain disturbances. \\ Sky radiance might be influenced by multiple reflectance between the sea surface albedo and the atmosphere and in general, sky radiance increases rapidly while viewing zenith angle is near to $90\,^{\circ}$ (i.e. in the evening hours). Therefore, ideally measurements should be done at noon, with as low zenith angle (Fig.\ref{fig:A.2-1})\label{zenith} as possible; otherwise, if the solar zenith angle is too high (e.g. approaching $60-70\,^{\circ}$), received data may contain noise. For \ac{WASI} simulations reflection factor of sky radiance is taken as 0.0201 with simulated ideal conditions: viewing angle ($0\,^{\circ}$ = nadir). The interval of water temperature taken as default lies in the diapason $17-25\,^{\circ}$ as to simulate the conditions of the Mediterranean Sea. However, in real time conditions the sun zenith angle of $45\,^{\circ}$ has been accepted as suitable.\vspace{1ex}\\
The anisotropy factor of upwelling radiation or the quality (Q-) factor, showing the directionally dependency of the radiance, is taken as 5. We accept some values of optical parameters as default values at \ac{WASI}. Thus, concentration of phytoplankton is accepted at the interval of $0.035-0.089 mg^{-1}$ and concentration of large suspended particles is given to 8. Reference wavelength for \ac{CDOM} (Gelbstoff) absorption is equal to 440. The coefficient of attenuation remains equal to 1.0546, as set up by default at \ac{WASI}.\vspace{1ex}\\
We also assume that concentration of non-chlorophyll particles (absorption at 0) as well as concentration of small suspended particles is equal to zero, so we do not count them in this work. Exponent of \ac{CDOM} (Gelbstoff) absorption is accepted as 0.0140. Finally, the \ac{BRDF} of bottom reflectance (sand), which defines the reflection of light  is at an opaque surface (Fig.A.2.5), is assumed to be 0.318 $sr^{-1}$.

%%%%%%%%%%%%%%%%%  Section2   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section[Overview of literature]{Seagrass monitoring: overview of literature and research resources} \label{sec:2}
\markright{II. Seagrass monitoring: overview of literature and research resources}
\renewcommand{\headrulewidth}{0.4pt}
\counterwithin{figure}{section}
\counterwithin{table}{section}

@@

\subsection[Measuring water optical properties]{Measuring water optical properties: \\hyperspectral radiometers}\label{sec:2-2}\label{page-12}
Among various research approaches for the seagrass mapping the most popular methods are based on application of the remote sensing techniques,
 enabling to detect and discriminate the substratum and various vegetation types growing on the seafloor.
The application of the remote sensing data for seagrass mapping is based on our knowledge of the
spectral reflectance properties of the target objects, and using it for the classification of these objects
on the image. In case of seagrasses it is spectral reflectance of the seafloor cover types, which can be
analyzed using measurements of optical properties of sea water: radiance and irradiance.
Optical remote sensing methods can get through the clear waters to approximately 15–30 m \cite{Mumby04}\label{Mumby04}.
When sunlight enters the waters and goes down into the water column, parts of the
electromagnetic energy are absorbed and scattered, which is determined by the optical and physical
properties of the water, e.g. concentration of suspended particles, chlorofill, colored dissolved
organic matter (gelbstoff) that make up the water content \cite{Ackleson2003}\label{Ackleson, 2003}. Besides, light is strongly
dependent on wavelengths, i.e. it is greater in blue wavelengths (400 nm) than in others.

@@

\subsection[Radiative Transfer Models (RTM)]{Radiative Transfer Models (RTM) for the simulation of water optical properties: a brief review of existing tools}\label{sec:2-3}

@@

\subsection[Application of the remote sensing...]{Application of the remote sensing data towards seagrass mapping}\index[concepts]{Remote sensing}\label{page-15}
Various methods and approaches have been applied towards mapping of the seagrasses, based on
digitized aerial photographs, \ac{GPS} data, remote sensing and \ac{SCUBA}-based (Fig.\ref{fig:A.52}) fieldwork measurements.
\ac{SCUBA}-based \textit{in-situ} observations, though providing high resolution and accuracy results in seagrass
mapping, is limited in application, because of their time consumption, weather-dependency and
unsuitability for the case of monitoring large areas of water for small-scale mapping. 
The underwater videography\index[concepts]{Remote sensing!underwater videography} with a \ac{GPS}\index[concepts]{Remote sensing!GPS} is a tool of seagrass monitoring which has certain advantages, i.e.
high spatial and visual resolution, non-destructive sampling, effectiveness at all depths and rapid data
collection in the field \cite{Schultz08}\label{Schultz08}.

@@

\subsubsection[Review of multispectral imagery...]{Review of multispectral imagery used for seagrass mapping}\label{sec:2.4.1}\index[concepts]{Remote sensing!multispectral imagery} 
Seagrass mapping using remotely sensed data\label{page-18} from multispectral sensors is based on the classification
and discrimination of the seafloor types using their spectral characteristics in different wavebands.

@@

\subsubsection[Review of hyperspectral imagery...]{Review of hyperspectral imagery used for seagrass mapping}\label{sec:2.4.2}\index[concepts]{Remote sensing!hyperspectral imagery}
The application of the hyperspectral sensors \label{page-20} is most effective and provide more accurate classification
results  (Fig.\ref{fig:2.5}), due to their higher spectral resolution \cite{Bharathi03}\label{Bharathi03} with interval
narrows to 10 nanometers, while broadband sensors are limited to the spectral width of ca
150 nm  (Fig.\ref{fig:2.3}). 
Hyperspectral imagery is acquired through the simultaneous acquisition of images in many
narrow, contiguous spectral bands from hyperspectral scanners (mostly) cover the 400-
to 2500-nm spectral bands \cite{Schmidt03}\label{Schmidt03}.
Perhaps, the most advantageous and general characteristics of hyperspectral imagery is its
high spectral resolution, desirable for the case of seagrass monitoring. 
The most suitable scanner fit for detailed seagrass mapping would cover bands of 550-750 nm and have a spectrum
resolution of 5-15 nm \cite{Fyfe04}\label{Fyfe04}, which exactly characterizes typical airborne and
hyperspectral satellite scanners. Important features may be detected in the narrow
wavelengths of hyperspectral imagery, while this information can be lost in the broader
wavelengths of other sensors. With its 126 spectral bands, \ac{HyMap} imagery enables to
distinguish features of interest, i.e. seagrass types \cite{Peneva07}\label{Peneva07}, which is the major
advantage of the of hyperspectral data for mapping landscapes of the seagrasses.
While comparing multispectral imagery with airborne hyperspectral, the last one showed higher
overall accuracies \cite{Phinn08}\label{Phinn08}. Several, regular and narrow (10 nm) spectral bands\label{sec:2.4.2}\index[concepts]{Remote sensing!hyperspectral imagery!spectral bands}, specific for
hyperspectral imagery, are strong tool enabling to detect even slight and subtle differences in the
spectral reflectance between various seafloor types, e.g. different seagrass species at diverse depths,
algae, corals, dark-coloured sands or other types of sediments \cite{Hochberg03a}\label{Hochberg03a}. 
Therefore, there is great potential of the application of hyperspectral remote sensing imagery towards seagrass
mapping at species level, as long as they are distinguishable spectrally (Fig.\ref{fig:2.6}) which, for example, has been
tested in the case study of Australian marine ecosystems by Fyfe \cite{Fyfe04}\label{Fyfe04}. 
Application of various classifications methods, \textit{inter alia} maximum likelihood, minimum distance and means, towards
hyperspectral imagery \cite{Peneva08}\label{Peneva08}, combined with fieldwork measurements ensures accurate
mapping results with the maximum likelihood methods producing the best results. 
Therefore, accurate mapping of the seagrass landscapes and other seafloor types using remote sensing
approaches requires application of high-spatial resolution (higher than 5 m) or hyperspectral imagery.
Comparative analysis of the application of hyperspectral\label{sec:2.4.2}\index[concepts]{Remote sensing!hyperspectral imagery!application} and multispectral imagery towards the
seafloor types classification \cite{Hochberg03a}\label{Hochberg03a} demonstrates that coral, seagrasses and
sand very well distinguishable in their spectra with an overall classification accuracy of 98 \%.
However, the use of data from various sensors, both hyperspectral and multispectral, is possible and
reasonable, as soon as it meets the research specific objective. 
Thus, the use of multispectral imagery
with high spatial resolution is preferable to using hyperspectral medium resolution data in case of
mapping benthic vegetation in areas where the spatial heterogeneity is very high \cite{Vahtmae07}\label{Vahtmae07}.

\subsubsection{Comparison of seagrass spectra }\label{sec:2.5.3}

@@

The videographic measurements and photos, enabling to detect types of the
seafloor, were captured by means of three underwater digital video cameras \ac{Olympus ST} 8000  (Fig.\ref{fig:3.3d})\label{page-26}, at
five minutes interval along the tracklog path (Fig.\ref{fig:3.6}).
research sampling design also included measurements of depths (Fig. \ref{fig:A.4}), because bathymetry is one of the
most determining factors for the seagrass locations. Measurements of depths were performed during
the fieldwork using cord, \ac{iPAQ}, and markers in order to assure that the videometric measurements
are taken at depths not more than four meters, in the shelf area.
As a result, nine transects were established of one m wide and 20 m long track, to cover seagrass beds
with videographic measurements. The \ac{GPS} allowed to capture measurement locations on the \ac{iPAQ},
encapsulated in a plastic waterproof Otterbox (Fig \ref{fig:3.3b}). The camera were adjusted horizontally by a leveler and
mounted under the bottom of the boat Fig.\ref{fig:A.51}\label{page-27} to capture video footage and imagery (Fig.\ref{fig:3.5}). The data
were taken at proper weather conditions: sunny, serene and cloud-free days with glassy sea state. 
The locations of the route were randomly selected in the areas of the Ligaria beach, to ensure most dense
coverage of the seagrass meadows in the research area. The underwater measurements of the seagrass
coverage were carried out by taking video footage and photos of ca 0.5 m2 size each (Fig.\ref{fig:A.7}, Fig.\ref{fig:A.55}).
The results of the underwater videometric measurements include series of digital images helping to
classify seafloor cover types (Fig.\ref{fig:A.56}) and seagrass meadows, according to the differences in the structure,
colour, texture and shapes of the depicted objects\label{page-27}.

@@

%
\subsection{Analysis of spectral signatures}
The distinguishing spectral signatures for various seafloor types (e.g. seagrass species, coral reefs,
various types of sand, mud, other sediments, (Fig.\ref{fig:4.4})\label{page-31} exist in well-defined and narrow (10–20 nm)
wavelength ranges. 
The values of their spectral reflectance are accepted as constant. The results of
spectral measurements enable to analyze, whether \textit{P.oceanica} is spectrally distinct from other sea
floor types with changing environmental conditions, using the differences in their spectral signatures
on the graphs in a \ac{WASI} software. \\The \ac{WASI}, a
software tool for analyzing and simulating the most common types of spectra \cite{Gege05}\label{Gege05}, is highly
suitable for the seagrass spectral analysis.
There are several environmental characteristics, included in \ac{WASI} interface (Fig.\ref{fig:4.7}), which influence the
results of water spectral reflectance, e.g. different bottom depths, concentration of suspended particles
in water column, water temperature, sun angle, concentration of Gelbstoff (\ac{CDOM}), concentration of phytoplankton, aerosol scattering, exponent of backscattering by small
particle \cite{Gege04}\label{Gege04}.

>>>>

\documentclass[11pt]{article}
\special{papersize=140mm,210mm}
\usepackage{authblk}
\usepackage[english]{babel}
%\usepackage{authorindex}
\usepackage{color}
\usepackage[stable]{footmisc }
\usepackage{setspace}
\usepackage[printonlyused,withpage]{acronym}
\usepackage[left=30mm,right=20mm,top=30mm,bottom=35mm]{geometry}
\usepackage{pslatex}
%\usepackage{times}
%\usepackage{psnfss}
%\usepackage{glossary}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[numbers]{natbib}
\usepackage{calc}
\usepackage{amsmath}
\usepackage{longtable}
\usepackage[svgnames,table]{xcolor}
\usepackage[subfigure]{tocloft}
%\usepackage[table]{xcolor}
\usepackage{paralist}
%\usepackage{makeidx}\makeindex
%\usepackage[makeindex,protected]{splitidx}
%\usepackage{index}
\usepackage{imakeidx}
\makeindex[name=concepts,title=Index of Concepts,intoc]
\makeindex[name=people,title=Index of People,intoc,columns=3,columnseprule]
\usepackage[fit]{truncate}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[RO,LE]{\truncate{.95\headwidth}{\leftmark}}
\usepackage{fancyheadings}
\topmargin = 1pt
\headheight=10mm
\headsep=7pt
%\ensuremath
\usepackage{lastpage}
\usepackage{float}
\usepackage{subfig}
\usepackage{caption}
\usepackage{chngcntr}
\counterwithin{figure}{section}
\counterwithin{table}{section}
\usepackage{appendix}
\captionsetup{margin=10pt,font=small,labelfont=bf}
\usepackage{wrapfig}
\definecolor{darkblue}{rgb}{0.0,0.0,0.3}
\setcounter{tocdepth}{4}
\parindent 0pt
%\floatstyle{boxed} 
\restylefloat{figure}
\usepackage{hyperref}
\hypersetup{pdftitle={Seagrass mapping and monitoring along the coasts of Crete, Greece}, pdfauthor={Lemenkova Polina}, pdfsubject={MScThesis}, pdfcreator={Lemenkova Polina}, pdfproducer={Lemenkova Polina}, colorlinks=true,linkcolor=Navy, citecolor=Crimson, filecolor=magenta, urlcolor=blue,pdfkeywords={P.oceanica, seagrass, spectral reflectance, remote sensing, WASI, RTM, Google Earth, mapping, Crete}}

@@

\subsection{Research questions}\index[concepts]{Research!questions}
	\begin{itemize}	
		\item Is \textit{P.oceanica} spectrally distinct from carbonate sand with varying \textit{in-situ} conditions ?
		\item Do broadband and hyperspectral sensors provide enough radiometric information for spectral discrimination of seagrass, and therefore, can be used for mapping of \textit{P.oceanica} ?
	\end{itemize}

\subsection{Hypotheses}\index[concepts]{Research!questions!hypothesis}
A statistical testing will be used to compare between the spectral responses of the different seafloor cover types (i.e. sand and \textit{P. oceanica}),
 whether it is spectrally distinct and at least one pair is statistically different at every spectral
band.\vspace{1ex}\\
For the \emph{Research Question 1} the Hypothesis $H_o$ claims: seagrass types are not spectrally distinct from other seafloor types with
varying \textit{in-situ} conditions, which means $H_o: \mu1 =\mu2 =\mu3=...= \mu n$.
The alternative Hypothesis $H_a$ claims the opposite statement: seagrass is spectrally
distinct with varying \textit{in-situ} conditions, $H_o:\mu1\neq\mu2\neq\mu3\neq...\neq\mu n$.\vspace{1ex}\\
For the \emph{Research Question 2} the Hypothesis $H_o$ claims: broadband and hyperspectral remote sensing data can be used for the mapping of \textit{P.oceanica}, because they do not provide enough radiometric information to discriminate sand from seagrass, which means $H_o:\mu1=\mu2=\mu3=...=\mu n$. The alternative Hypothesis $H_a$ claims the opposite statement: broadband and hyperspectral sensors do prove to provide enough radiometric information to discriminate sand from seagrass, 
$H_o: \mu1\neq\mu2\neq\mu3\neq ... \neq\mu n$.\vspace{1ex}\\
The distribution of the spectral responses at every spectral band is assumed to be normal as well as
the equality of the statistical variances.
The hypothesis testing is suggested to be carried out using the \ac{ANOVA} statistical test. The purpose of
\ac{ANOVA} test is to visualize in an effective and quick way the spectral differences between seagrass
species and their spatial distribution. Thus, the key hypotheses of the research will be tested to prove
whether the results of the research are accurate, reasonable and correct.

\subsection{Assumptions}\index[concepts]{Research!assumptions}
The general research assumption, used in this work, in order to make feasible application of the remote sensing methods, includes some statements about the \textit{in-situ} atmospheric conditions, viewing angle, wave backscattering and other optical properties of the environment. They all definitely play a certain role and impact final results of optical measurements, but for practical reasons we have chosen to ignore their contribution in the spectral separability of seafloor cover types. We briefly list below general research assumptions for the optical properties of the environmental variables used in this study. \vspace{1ex}\\
Weather conditions for the measurements are assumed to be ideal: clear, sunny, windless day. Otherwise wind roughens water surface causing sun glitter, and values of spectral reflectance may contain disturbances. \\ Sky radiance might be influenced by multiple reflectance between the sea surface albedo and the atmosphere and in general, sky radiance increases rapidly while viewing zenith angle is near to $90\,^{\circ}$ (i.e. in the evening hours). Therefore, ideally measurements should be done at noon, with as low zenith angle (Fig.\ref{fig:A.2-1})\label{zenith} as possible; otherwise, if the solar zenith angle is too high (e.g. approaching $60-70\,^{\circ}$), received data may contain noise. For \ac{WASI} simulations reflection factor of sky radiance is taken as 0.0201 with simulated ideal conditions: viewing angle ($0\,^{\circ}$ = nadir). The interval of water temperature taken as default lies in the diapason $17-25\,^{\circ}$ as to simulate the conditions of the Mediterranean Sea. However, in real time conditions the sun zenith angle of $45\,^{\circ}$ has been accepted as suitable. 
The accepted model-specific parameters of the \ac{WASI} simulator are shown in the Table \ref{tab:3}.\label{page-10} \vspace{1ex}\\
The anisotropy factor of upwelling radiation or the quality (Q-) factor, showing the directionally dependency of the radiance, is taken as 5. We accepted some values of model-specific optical parameters as default values at \ac{WASI} simulator, which are shown in the Table \ref{tab:3}. Thus, concentration of phytoplankton is accepted at the interval of $0.035-0.089 mg^{-1}$ and concentration of large suspended particles is given to 8. Reference wavelength for \ac{CDOM}, or Gelbstoff, absorption is equal to 440. The backscattering (Fig.\ref{fig:backscat}) is accepted to be $0.00144m^{-1}$. The coefficient of attenuation remains equal to 1.0546, as set up by default at \ac{WASI}.\vspace{1ex}\\
We also assume that concentration of non-chlorophyll particles (absorption at 0) as well as concentration of small suspended particles is equal to zero, so we do not count them in this work. Exponent of \ac{CDOM} (Gelbstoff) absorption is accepted as 0.0140. Finally, the \ac{BRDF} of bottom reflectance (sand), which defines the reflection of light  is at an opaque surface (Fig.\ref{fig:BRDF})\label{page-11}, is assumed to be 0.318 $sr^{-1}$.

@@

The pre-processing of the images includes imagery corrections for atmospheric noises and effects of the water
column. Reflectance spectra of the seagrass canopy at different depths of the water-column are
analysed for the discrimination of their spectral signatures, enabling to separate various seafloor types
during classification. The results of the of imagery classification are analysed for the detection of the
dynamics in \textit{P.oceanica} seagrass distribution along the northern coasts of Crete. Aerial imagery from
Google Earth with high spatial resolution, suitable for the large-scale detailed mapping of seagrass
mattes, is used for the improvement of the accuracy of large seagrass meadows and separate mattes
within the meadows. The \textit{in situ} underwater videometric measurements of the seafloor are collected
during the fieldwork in Crete, for the validation of the classification results and to determine the exact
current distribution of the \textit{P.oceanica} meadows. The image processing includes steps of the remote
sensing techniques, i.e. calibration, masking from land and cloud, atmospheric correction, sea surface
glint and depth effects correction as recommended \cite{Matarrese08}\label{Matarrese08}. During the image classification
working step the training sites for the supervised classification methods are designed, as well as its
control and trials of different classification approaches (Unsupervised, K-means or Isodata;
Supervised, Maximum Likelihood).
\pagebreak

%%%%%%%%%%%%%%%%%  Section2   %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section[Seagrass monitoring: overview of literature and research resources]{Seagrass monitoring: overview of literature and research resources} \label{sec:2}
\markright{Seagrass monitoring: overview of literature and research resources}
\renewcommand{\headrulewidth}{0.4pt}
\counterwithin{figure}{section}
\counterwithin{table}{section}

@@

\subsection[Measuring optical properties of benthic vegetation: hyperspectral radiometers]{Measuring optical properties of benthic vegetation: hyperspectral radiometers%
\sectionmark{Measuring optical properties}}
\sectionmark{Overview of literature...}\label{sec:2-2}\label{page-12}

Among various research approaches for the seagrass mapping the most popular methods are based on application of the remote sensing techniques,
 enabling to detect and discriminate the substratum and various vegetation types growing on the seafloor.
The application of the remote sensing data for seagrass mapping is based on our knowledge of the
spectral reflectance properties of the target objects, and using it for the classification of these objects
on the image. In case of seagrasses it is spectral reflectance of the seafloor cover types, which can be
analyzed using measurements of optical properties of sea water: radiance and irradiance.
Optical remote sensing methods can get through the clear waters to approximately 15–30 m \cite{Mumby04}\label{Mumby04}.
When sunlight enters the waters and goes down into the water column, parts of the
electromagnetic energy are absorbed and scattered, which is determined by the optical and physical
properties of the water, e.g. concentration of suspended particles, chlorofill, colored dissolved
organic matter (gelbstoff) that make up the water content \cite{Ackleson2003}\label{Ackleson, 2003}. Besides, light is strongly
dependent on wavelengths, i.e. it is greater in blue wavelengths (400 nm) than in others.

@@

%\subsection[Radiative Transfer Models (RTM)]{Radiative Transfer Models (RTM) for the simulation of water optical properties: a brief review of existing software tools}\label{sec:2-3}

\subsection[Radiative Transfer Models (RTM) for the simulation of water optical properties: a brief review of existing software tools]{Radiative Transfer Models (RTM) for the simulation of water optical properties: a brief review of existing software tools%
\subsectionmark{Radiative Transfer Models (RTM)}}
\sectionmark{Overview of literature...}\label{sec:2-3}

@@

\subsection[Application of the remote sensing data towards seagrass mapping]{Application of the remote sensing data towards seagrass mapping}\index[concepts]{Remote sensing}\label{page-15}

Various methods and approaches have been applied towards mapping of the seagrasses, based on
digitized aerial photographs, \ac{GPS} data, remote sensing and \ac{SCUBA}-based (Fig.\ref{fig:A.52}) fieldwork measurements.
\ac{SCUBA}-based \textit{in-situ} observations, though providing high resolution and accuracy results in seagrass
mapping, is limited in application, because of their time consumption, weather-dependency and
unsuitability for the case of monitoring large areas of water for small-scale mapping. 
The underwater videography\index[concepts]{Remote sensing!underwater videography} with a \ac{GPS}\index[concepts]{Remote sensing!GPS} is a tool of seagrass monitoring which has certain advantages, i.e.
high spatial and visual resolution, non-destructive sampling, effectiveness at all depths and rapid data
collection in the field \cite{Schultz08}\label{Schultz08}.

@@

\subsubsection[Review of multispectral imagery used for seagrass mapping]{Review of multispectral imagery used for seagrass mapping}\subsectionmark{Review of multispectral imagery used for seagrass mapping}\label{sec:2.4.1}\index[concepts]{Remote sensing!multispectral imagery}

Seagrass mapping using remotely sensed data\label{page-18} from multispectral sensors is based on the classification
and discrimination of the seafloor types using their spectral characteristics in different wavebands.

@@

\subsubsection[Review of hyperspectral imagery used for seagrass mapping]{Review of hyperspectral imagery used for seagrass mapping}\subsectionmark{Review of hyperspectral imagery used for seagrass mapping}\label{sec:2.4.2}\index[concepts]{Remote sensing!hyperspectral imagery}

The application of the hyperspectral sensors \label{page-20} is most effective and provide more accurate classification
results  (Fig.\ref{fig:2.5}), due to their higher spectral resolution \cite{Bharathi03}\label{Bharathi03} with interval
narrows to 10 nanometers, while broadband sensors are limited to the spectral width of ca
150 nm  (Fig.\ref{fig:2.3}). 
Hyperspectral imagery is acquired through the simultaneous acquisition of images in many
narrow, contiguous spectral bands from hyperspectral scanners (mostly) cover the 400-
to 2500-nm spectral bands \cite{Schmidt03}\label{Schmidt03}.
Perhaps, the most advantageous and general characteristics of hyperspectral imagery is its
high spectral resolution, desirable for the case of seagrass monitoring. 
The most suitable scanner fit for detailed seagrass mapping would cover bands of 550-750 nm and have a spectrum
resolution of 5-15 nm \cite{Fyfe04}\label{Fyfe04}, which exactly characterizes typical airborne and
hyperspectral satellite scanners. Important features may be detected in the narrow
wavelengths of hyperspectral imagery, while this information can be lost in the broader
wavelengths of other sensors. With its 126 spectral bands, \ac{HyMap} imagery enables to
distinguish features of interest, i.e. seagrass types \cite{Peneva07}\label{Peneva07}, which is the major
advantage of the of hyperspectral data for mapping landscapes of the seagrasses.
While comparing multispectral imagery with airborne hyperspectral, the last one showed higher
overall accuracies \cite{Phinn08}\label{Phinn08}. Several, regular and narrow (10 nm) spectral bands\label{sec:2.4.2}\index[concepts]{Remote sensing!hyperspectral imagery!spectral bands}, specific for
hyperspectral imagery, are strong tool enabling to detect even slight and subtle differences in the
spectral reflectance between various seafloor types, e.g. different seagrass species at diverse depths,
algae, corals, dark-coloured sands or other types of sediments \cite{Hochberg03a}\label{Hochberg03a}. 
Therefore, there is great potential of the application of hyperspectral remote sensing imagery towards seagrass
mapping at species level, as long as they are distinguishable spectrally (Fig.\ref{fig:2.6}) which, for example, has been
tested in the case study of Australian marine ecosystems by Fyfe \cite{Fyfe04}\label{Fyfe04}. 
Application of various classifications methods, \textit{inter alia} maximum likelihood, minimum distance and means, towards
hyperspectral imagery \cite{Peneva08}\label{Peneva08}, combined with fieldwork measurements ensures accurate
mapping results with the maximum likelihood methods producing the best results. 
Therefore, accurate mapping of the seagrass landscapes and other seafloor types using remote sensing
approaches requires application of high-spatial resolution (higher than 5 m) or hyperspectral imagery.
Comparative analysis of the application of hyperspectral\label{sec:2.4.2}\index[concepts]{Remote sensing!hyperspectral imagery!application} and multispectral imagery towards the
seafloor types classification \cite{Hochberg03a}\label{Hochberg03a} demonstrates that coral, seagrasses and
sand very well distinguishable in their spectra with an overall classification accuracy of 98 \%.
However, the use of data from various sensors, both hyperspectral and multispectral, is possible and
reasonable, as soon as it meets the research specific objective. 
Thus, the use of multispectral imagery
with high spatial resolution is preferable to using hyperspectral medium resolution data in case of
mapping benthic vegetation in areas where the spatial heterogeneity is very high \cite{Vahtmae07}\label{Vahtmae07}.

\subsubsection[Comparison of seagrass spectra]{Comparison of seagrass spectra}\subsectionmark{Comparison of seagrass spectra}\label{sec:2.5.3}

@@

The videographic measurements and photos, enabling to detect types of the
seafloor, were captured by means of three underwater digital video cameras \ac{Olympus ST} 8000  (Fig.\ref{fig:3.3d})\label{page-26}, at
five minutes interval along the tracklog path (Fig.\ref{fig:3.6}).
research sampling design also included measurements of depths (Fig. \ref{fig:A.4}), because bathymetry is one of the
most determining factors for the seagrass locations. Measurements of depths were performed during
the fieldwork using cord, \ac{iPAQ}, and markers in order to assure that the videometric measurements
are taken at depths not more than four meters, in the shelf area.
As a result, nine transects were established of one m wide and 20 m long track, to cover seagrass beds
with videographic measurements. The \ac{GPS} allowed to capture measurement locations on the \ac{iPAQ},
encapsulated in a plastic waterproof Otterbox (Fig \ref{fig:3.3b}). The camera were adjusted horizontally by a leveler and
mounted under the bottom of the boat Fig.\ref{fig:A.51}\label{page-27} to capture video footage and imagery (Fig.\ref{fig:3.5}). The data
were taken at proper weather conditions: sunny, serene and cloud-free days with glassy sea state. 
The locations of the route were randomly selected in the areas of the Ligaria beach, to ensure most dense
coverage of the seagrass meadows in the research area. The underwater measurements of the seagrass
coverage were carried out by taking video footage and photos of ca 0.5 $m^2$ size each (Fig.\ref{fig:A.7}, Fig.\ref{fig:A.55}).
The results of the underwater videometric measurements include series of digital images helping to
classify seafloor cover types (Fig.\ref{fig:A.56}) and seagrass meadows, according to the differences in the structure,
colour, texture and shapes of the depicted objects\label{page-27}.

@@

%
\subsection{Analysis of spectral signatures}
The distinguishing spectral signatures for various seafloor types (e.g. seagrass species, coral reefs,
various types of sand, mud, other sediments, (Fig.\ref{fig:4.4})\label{page-31} exist in well-defined and narrow (10–20 nm)
wavelength ranges. 
The values of their spectral reflectance are accepted as constant. The results of
spectral measurements enable to analyze, whether \textit{P.oceanica} is spectrally distinct from other sea
floor types with changing environmental conditions, using the differences in their spectral signatures
on the graphs in a \ac{WASI} software. \\The \ac{WASI}, a
software tool for analyzing and simulating the most common types of spectra \cite{Gege05}\label{Gege05}, is highly
suitable for the seagrass spectral analysis.
There are several environmental characteristics, included in \ac{WASI} interface (Fig.\ref{fig:4.7}), which influence the
results of water spectral reflectance, e.g. different bottom depths, concentration of suspended particles
in water column, water temperature, sun angle, concentration of Gelbstoff (\ac{CDOM}), concentration of phytoplankton, aerosol scattering, exponent of backscattering by small particles \cite{Gege04}\label{Gege04}.
